A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)

### 2018-04-15 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
For First Submission 15th April 2018

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a SVD-based method was developed to reduce the order of porous media permeability field (\ie upscale). This method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits performance. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the SVD upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

\medskip
Finally, $\mathbf{K}\left(\mathbf{r}\right)$, the focus of this work, is the absolute permeability tensor of the porous medium and it is a function of the vector position $\mathbf{r}$. $\mathbf{K}$ is prescribed in the system of governing conservative equations, \ie it is an indepedent input variable. Since $\mathbf{K}$ is petrophysical rock property that varies spatially, it is not possible to determine at every spatial coordinate. And, even if this information is available at every spatial coordinate, it would require immense computer resources to process during flow simulations. This is the reason upscaling is required for processing permeability variables. Upscaling can be performed from the continuum scale, across the micro scale ($mm$), local scale and meso scale ($m$) to the field scale ($km$)\cite{ECMI_2004}. The upscaling discussed here is from the micro scale also known as the Darcy scale.

@@

The model order reduction for the permeability field discussed here is concerned with reducing the volume of permeability data within the permeability space. Consider for instance a small square which is discretized to contain four squares such that $4$ data values are defined in the middle of the four squares. The permeability space of $4$ can be reduced to a smaller value of $1$ through an averaging process such as arithmetic, harmonic or geometric mean. This type of reduction is the focus of this paper, albeit the main focus of this paper is achieving this reduction through SVD in the principal component spaces.

@@

The second step involves interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) to reduce the data in each column of  $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

@@

The BaseCase permeability model used here retains the checkerboard desgined used by the recently named authors. Albeit, the permeability data for each block in the set of diagonally opposite blocks have similar value (\ie not exactly the same as in the papers referred to), although one set can still be referred to as a set of high permeability blocks while the other set is referred to as a set of low permeability blocks. Furthermore, within each block, the permeability field varies within a pre-defined range and consists of $1600$ permeability values. More details on the design of the permeability field for the BaseCase is available in section~\ref{subsubsection:basecase_preprocess_algorithm}. Figures~\ref{fig:PermFields}a and~\ref{fig:PermFields}b give a pictorial representation of the permeability field for the base case with the mesh discretization visible and invisible, respectively.  Using the labelling scheme used by \citet{dawe_2008}, the quadrant model block $K1,~K2,~K3$ and $K4$ are top left, bottom left, top right and bottom right, respectively.

@@

\subsubsection{BaseCase Pre-processing Algorithm}\label{subsubsection:basecase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item A function is used to randomly generate uniform (\ie Gaussian) data for each block. The range for each block is as below:
  \begin{enumerate}[a)]
    \item \textbf{K1} : $700 - 900$ mD
    \item \textbf{K2} : $50 - 150$ mD
    \item \textbf{K3} : $100 - 200$ mD
    \item \textbf{K4} : $1000 - 1200$ mD
  \end{enumerate}                                                    
  \item The number of data points generated for each block is $1600$ and the data is randomly allocated within each block in a structured pattern. Albeit, the simulation mesh is unstructured.
  \item Merge data from different blocks into a single file retaining the original ``structured pattern'' (in step 1 above) for the randomly positioned data in each block. This file will be used by the CVFEM-based flow simulator to prescribe the permeability field for the base case.
\end{enumerate}

\subsubsection{PDFCase Pre-processing Algorithm}\label{subsubsection:pdfcase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item Data from step one of the pre-processing algorithm for the base case (Section~\ref{subsubsection:basecase_preprocess_algorithm}) is used to calculate mean and standard deviation for each block of the permeability field.
  \item A function which generates a Gaussian data set using arithmetic mean and standard deviation is used to stochastically produce a quarter of data points from the BaseCase for each of the four blocks.
  \item The number of data points \red{(PREVIOUSLY, IT WAS ``THE NUMBER OF DATA'' ... HOWEVER DATA IS NOT AN ACCOUNTABLE NOUM, THEREFORE YOU MAY USE EITHER ``AMOUNT OF DATA'' OR ``NUMBER OF DATA POINTS'' ... I REPLACED MOST OF ENTRIES -RE DATA AS AN ACCOUNTABLE NOUM ... PLEASE REVISE THE MANUSCRIPT ACCORDINGLY.)} generated for each block is $400$ and the data is randomly positioned within each block in a structured pattern. As in the base case, the simulation grid is unstructured.
  \item Finally, the structured data from each block is combined into a single file and then used by the flow simulator, which merge the permeability mapping to the mesh with the appropriate element-pairs.
\end{enumerate}

\subsubsection{SVDCase Pre-processing Algorithm}\label{subsubsection:svdcase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item The permeability field obtained in step 3 of the pre-processing algorithm for the base case is factorized using SVD to obtain $\mathbf{U}, \Sigma,$ and $\mathbf{V}$.
  \item A selected number ($z$) of singular values are retained in decreasing order of magnitude to form a new diagonal matrix, $\Sigma_{new}$.
  \item The first $z$ columns of $\mathbf{U}$ and $\mathbf{V}$ are selected to form new truncated matrices call $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$ respectively.
  \item Within each column of $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$, a linear interpolation is performed to form a new column with the number of data in each column reduced to half of the original column. These new columns are recombined to form $\mathbf{U_{new2}}$ and $\mathbf{V_{new2}}$.
  \item The permeability field used for the SVD case is the matrix $\mathbf{A_{new2}}$, which is defined as the product of $\mathbf{U_{new2}} \mathbf{\Sigma_{new}} \mathbf{V_{new2}^{\intercal}}$. It is a permeability field with a quarter of the data in the base case and therefore a permeability field with reduced order.  
\end{enumerate}

@@

Mesh resolution for the the BaseCase and upscaled cases are shown in Fig,~\ref{fig:HiRes_LowRes_Mesh}. Domain of the BaseCase (\ie high-resolution mesh) was discretised with $4112$ \PNDG[1]{1} \red{(IS THIS TRUE?)} triangular element-pairs, whilst in upscaled cases the domain was discretised with $728$ \PNDG[1]{1} \red{(IS THIS TRUE?)} triangular element-pairs. Therefore the resolution for the upscaled cases is reduced by a factor of approximately $5.6$.

Figure~\ref{fig:PermFields} compares the permeability distribution for all models. \red{Permeability legend in Fig.~\ref{fig:PermFields}(a) is representative for all the other figures (\ie Fig.~\ref{fig:PermFields}(b) to (f))} \red{(THIS INFO SHOULD BE IN THE CAPTION OF THE FIGURES INSTEAD)}. The cases are shown without the edges for elements which discretize the domain in Fig.~\ref{fig:PermFields}(b) to (f).

@@

With respect to the PDFCase and for this specific resolution, it is important to note that Fig.~\ref{fig:PermFields}(e) is only one realiztion of infinitely many possible representations of the PDFCase. Meanwhile, the SVDCase will retain this specific (\ie Fig.~\ref{fig:PermFields}(e)) permeability field for this specific resolution (\ie if linear interpolation is used, but this digresses from the main point and will be discussed in future work). Therefore, the permeability value at a specific point in the PDFCase is not directly related to the data value at the same point in the BaseCase but the SVDCase has this direct connection through the interpolation \red{(SENTENCE IS SLIGHTLY CONFUSING ... NOT SURE WHAT YOU WANT TO CONVEY)}.

@@

\medskip
\red{(I GUESS YOU SHOULD USE ONE PARAGRAPH JUST TO DESCRIBE (TO THE READER WHO HASNT READ THE OTHER REFERENCES) THE PHYSICS ASSOCIATED WITH THE BASECASE -- FIG 4 LINKING WITH THE REFERENCES ... REMEMBER THIS IS YOUR WORK'S VALIDATION!!! THIS BENCHMARK WOULD: (A) IF THE SIMULATION WAS PERFORMED W HOMOGENEOUS MEDIA (AT EACH BLOCK) A PISTON SHAPE BEHAVIOR (AT EACH BLOCK, I.E., A BLOCK DIVIDED BUCKLEY-LEVERETT BEHAVIOR) COULD BE NOTICED HOWEVER AS THE WHOLE SYSTEM IS HETEROGENEOUS THIS MAY NOT BE CLEARLY NOTICED; (B) CROSS-FLOW, I.E., PREFERENTIAL FLOW  THROUGH REGIONS OF HIGH PERMEABILITY; (C) FLOW QUALITATIVE BEHAVIOUR MATCHES THE EXPERIMENTS  ETC ETC ETC MAKE IT STRONG !)}

@@

Figure~\ref{fig:Saturationfield@t=1.15s} compares phase 1 saturation distribution for all the cases at $t=1.15s$. Generally, saturation distribution for all cases closely resemble each other with arithmetic and harmonic mean (Figs.~\ref{fig:Saturationfield@t=1.15s}(b) and (c)) being the most identical. Albeit, closer inspection will reveal that there is more injection in Fig.~\ref{fig:Saturationfield@t=1.15s}(b) compared to (c)\red{(WHAT DO YOU MEAN W ``MORE INJECTION''? MASS FLOW RATE OF INJECTION (BC) WAS THE SAME FOR ALL CASES ...DO YOU MEAN THE DISPLACEMENT IN THE RHS QUADRANTS?)}. This is logical giving that the block mean values calculated for arithmetic mean is higher than that for harmonic mean. However, the important feature of Figs.~\ref{fig:Saturationfield@t=1.15s}(b) and (c) is that they poorly represent the BaseCase (Fig.~\ref{fig:Saturationfield@t=1.15s}(a)) at the top left hand corner of block $K2$.

@@

Results show that the permeability field data reduction can be specifically defined for the PDFCase and the SVDCase but not for the ArithMeanCase and the HarmMeanCase. This is because the latter cases will produce the same permeability field irrespective of the upgrid resolution. The SVDCase is the only technique that does not require a prior knowledge of blocks with similar permeability values before upscaling.

>>>>

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flows in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domains. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all available data values. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a SVD-based method was developed to reduce the order of porous media permeability field (\ie upscale). This method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits/performance. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the SVD upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

\medskip
Finally, $\mathbf{K}\left(\mathbf{r}\right)$, the focus of this work, is the absolute permeability tensor of the porous medium and it is a function of the vector position $\mathbf{r}$. $\mathbf{K}$ is prescribed in the system of governing conservative equations, \ie it is an indepedent input variable. Since $\mathbf{K}$ is a petrophysical rock property that varies spatially, it cannot be determined at every spatial coordinate. And, even if this information is available at every spatial coordinate, it would require immense computer resources to process during flow simulations. This is the reason upscaling is required for processing permeability variables. Upscaling can be performed from the continuum scale, across the micro scale ($mm$), local scale and meso scale ($m$) to the field scale ($km$)\cite{ECMI_2004}. The upscaling discussed here is from the micro scale also known as the Darcy scale.

@@

The model order reduction for the permeability field discussed here is concerned with reducing the volume of permeability data within the permeability space. Consider for instance a small square which is discretized to contain four squares such that $4$ data points are defined in the middle of the four squares. The permeability space of $4$ can be reduced to a smaller value of $1$ through an averaging process such as arithmetic, harmonic or geometric mean. This type of reduction is the focus of this paper, albeit the main focus of this paper is achieving this reduction through SVD in the principal component spaces.

@@

The second step involves interpolation within each singular vector (column in $\mathbf{U}$ and row in $\mathbf{V^{\intercal}}$) to reduce the data points in each column of  $\mathbf{U}$ and in $\mathbf{V^{\intercal}}$. From this step,  $\mathbf{U_{new2}}$ and $\mathbf{V^{\intercal}_{new2}}$ are obtained. The reduced permeability field is calculated by multiplying  $\mathbf{U_{new2}}$, $\mathbf{\Sigma_{new}}$ and $\mathbf{V^{\intercal}_{new2}}$.

@@

The BaseCase permeability model used here retains the checkerboard desgined used by the recently named authors. Albeit, the permeability dataset for each block in the set of diagonally opposite blocks have similar value (\ie not exactly the same as in the papers referred to), although one set can still be referred to as a set of high permeability blocks while the other set is referred to as a set of low permeability blocks. Furthermore, within each block, the permeability field varies within a pre-defined range and consists of $1600$ permeability values. More details on the design of the permeability field for the BaseCase is available in section~\ref{subsubsection:basecase_preprocess_algorithm}. Figures~\ref{fig:PermFields}a and~\ref{fig:PermFields}b give a pictorial representation of the permeability field for the base case with the mesh discretization visible and invisible, respectively.  Using the labelling scheme used by \citet{dawe_2008}, the quadrant model block $K1,~K2,~K3$ and $K4$ are top left, bottom left, top right and bottom right, respectively.

@@

\subsubsection{BaseCase Pre-processing Algorithm}\label{subsubsection:basecase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item A function is used to randomly generate uniform (\ie Gaussian) dataset for each block. The range for each block is as below:
  \begin{enumerate}[a)]
    \item \textbf{K1} : $700 - 900$ mD
    \item \textbf{K2} : $50 - 150$ mD
    \item \textbf{K3} : $100 - 200$ mD
    \item \textbf{K4} : $1000 - 1200$ mD
  \end{enumerate}                                                    
  \item The number of data points generated for each block is $1600$ and the data values are randomly allocated within each block in a structured pattern. Albeit, the simulation mesh is unstructured.
  \item Merge the dataset from different blocks into a single file retaining the original ``structured pattern'' (in step 1 above) for the randomly positioned dataset for each block. This file will be used by the CVFEM-based flow simulator to prescribe the permeability field for the base case.
\end{enumerate}

\subsubsection{PDFCase Pre-processing Algorithm}\label{subsubsection:pdfcase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item The datasets from step one of the pre-processing algorithm for the base case (Section~\ref{subsubsection:basecase_preprocess_algorithm}) are used to calculate mean and standard deviation for each block of the permeability field.
  \item A function which generates a Gaussian dataset using arithmetic mean and standard deviation is used to stochastically produce a quarter of the data points from the BaseCase for each of the four blocks  (\ie $400$ data points for each block).
  \item The data values in the dataset for each block is randomly positioned within each block in a structured pattern. As in the base case, the simulation grid is unstructured.
  \item Finally, the structured/arranged dataset from each block is combined into a single file and then used by the flow simulator, which merge the permeability mapping to the mesh with the appropriate element-pairs.
\end{enumerate}

\subsubsection{SVDCase Pre-processing Algorithm}\label{subsubsection:svdcase_preprocess_algorithm}
\begin{enumerate}[1.]
  \item The permeability field obtained in step 3 of the pre-processing algorithm for the base case is factorized using SVD to obtain $\mathbf{U}, \Sigma,$ and $\mathbf{V}$.
  \item A selected number ($z$) of singular values are retained in decreasing order of magnitude to form a new diagonal matrix, $\Sigma_{new}$.
  \item The first $z$ columns of $\mathbf{U}$ and $\mathbf{V}$ are selected to form new truncated matrices call $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$ respectively.
  \item Within each column of $\mathbf{U_{new}}$ and $\mathbf{V_{new}}$, a linear interpolation is performed to form a new column with the number of data points in each column reduced to half of the original column. These new columns are recombined to form $\mathbf{U_{new2}}$ and $\mathbf{V_{new2}}$.
  \item The permeability field used for the SVD case is the matrix $\mathbf{A_{new2}}$, which is defined as the product of $\mathbf{U_{new2}} \mathbf{\Sigma_{new}} \mathbf{V_{new2}^{\intercal}}$. It is a permeability field with a quarter of the data values in the base case and therefore a permeability field with reduced order.  
\end{enumerate}

@@

Mesh resolution for the the BaseCase and upscaled cases are shown in Fig,~\ref{fig:HiRes_LowRes_Mesh}. Domain of the BaseCase (\ie high-resolution mesh) was discretised with $4112$ \PN[1]{1} triangular element-pairs, whilst in upscaled cases the domain was discretised with $728$ \PN[1]{1} triangular element-pairs. Therefore the resolution for the upscaled cases is reduced by a factor of approximately $5.6$.

Figure~\ref{fig:PermFields} compares the permeability distribution for all the models. As indicated in the figure's caption, the permeability legend in Fig.~\ref{fig:PermFields}(a) is representative for all the other figures (\ie Fig.~\ref{fig:PermFields}(b) to (f)). Additionally, Fig.~\ref{fig:PermFields}(b) to (f) show the cases without the element edges which discretize the domain.

@@

With respect to the PDFCase and for this specific resolution, it is important to note that Fig.~\ref{fig:PermFields}(e) is only one realiztion of infinitely many possible representations of the PDFCase. Meanwhile, the SVDCase will retain this specific (\ie Fig.~\ref{fig:PermFields}(e)) permeability field for this specific resolution (\ie if linear interpolation is used, but this digresses from the main point and will be discussed in future work). Therefore, the permeability value at a specific point in the PDFCase is not directly related to the data value at the nearest point in the BaseCase because the randomly generated PDF dataset is also randomly arranged within each block. However, the data value at a point in the SVDCase is directly related to the nearest data values with respect to the same point in the BaseCase, throught the interpolation which takes place in the PCS.

@@

Before going on to review the results of the simulations, it is beneficial to briefly consider the Buckley-Leverett and cross-flow behaviour which are be used to validate the results from this work.

\citet{buckley_1942} in 1942 derived the Buckley-Leverett equation, an analytical (albeit simplified) two phase flow equation. The equation determined that a plane of constant saturation progresses uniformly through the domain during the flow simulation. The equation is referred to as "simplified" because of the assumptions made while deriving it, some of which were ignored in this work. Principally, this work ignores the fact that the equation deals with injection into an homogeneous permeability field while this work deals injection into two blocks ($K1$ and $K2$) with different permeabilities. Additionally, the permeability field in each block is heterogeneous. However, the general validation test or expectation from Buckley-Leverett's influence on this work, is that, the injected fluid saturation in $K1$ and $K2$ is expected to have a piston-shape front within those blocks, subject some inconsistency at the boundary between the two blocks, and also within the blocks due to the heterogeneity.

\citet{dawe_2008} conducted experiments to investigate cross-flow, and as previously mentioned, the permeability field for the experimental work were somewhat replicated here for the simulations. Therefore, the work of \citet{dawe_2008} can be used to validate the resulting simulations from this work. In particular, it will be noted that the simulations matches the experiment at the boundary between $K1$ and $K2$. Additionally, the cross-flow from $K1$ to $K4$ is well represented in the simulations.

@@

Figure~\ref{fig:Saturationfield@t=1.15s} compares phase 1 saturation distribution for all the cases at $t=1.15s$. Generally, saturation distribution for all cases closely resemble each other with arithmetic and harmonic mean (Figs.~\ref{fig:Saturationfield@t=1.15s}(b) and (c)) being the most identical. Albeit, closer inspection will reveal that there is more displacement in Fig.~\ref{fig:Saturationfield@t=1.15s}(b) compared to (c). This is logical giving that the block mean values calculated for arithmetic mean is higher than that for harmonic mean. However, the important feature of Figs.~\ref{fig:Saturationfield@t=1.15s}(b) and (c) is that they poorly represent the BaseCase (Fig.~\ref{fig:Saturationfield@t=1.15s}(a)) at the top left hand corner of block $K2$.

@@

Results show that the permeability field data reduction can be specifically defined for the PDFCase and the SVDCase but not for the ArithMeanCase and the HarmMeanCase. This is because the latter cases will produce the same permeabiBuckleylity field irrespective of the upgrid resolution. The SVDCase is the only technique that does not require a prior knowledge of blocks with similar permeability values before upscaling.
